Notes on Control Plane Inference

### 2021-03-27 inference.tex
Outline proofs; Scope isn't managed correctly

% commands
\newcommand{\assert}{\mathop{\mathsf{assert}}}
\newcommand{\assume}{\mathop{\mathsf{assume}}}
\newcommand{\apply}{\mathsf{apply}}
\newcommand{\choiceop}{\rotatebox[origin=c]{90}{$\sqsubset\!\!\!\sqsupset$}}
\newcommand{\choice}{\mathbin{\choiceop}}
\DeclareMathOperator*{\bigchoice}{\scalerel*{\choiceop}{\sum}}
\renewcommand{\choose}[2]{\mathsf{choose}~#1~\mathsf{from}~#2}
\newcommand{\SKIP}{\mathsf{skip}}

@@

\begin{figure}[htp]
  \[\begin{array}{lcll}
    \multicolumn{3}{l}{\Value} \\
    v & ::= & [n]_w  & \textit{Bitvector Literal}\\
    \multicolumn{3}{l}{\SymbRow} \\
    \rho & ::= & \left\{
    \begin{array}{l}
      \matches = \List[x];\\
      \action = x;\\
      \data = \List[x]
    \end{array}\right\} & \textit{Symbolic Row}\\
    \multicolumn{3}{l}{\BVExpr} \\
    e & ::= & v & \textit{Value}\\
      & \mid & h.f & \textit{Header Access} \\
      & \mid & \rho.\matches[i] & \textit{Symbolic Match} \\
      & \mid & \rho.\action & \textit{Symbolic Action}\\
      & \mid & \rho.\data[i] & \textit{Symbolic Action Data}\\
      & \mid & x & \textit{Variable}\\
      & \mid & e \binop e & \textit{Arbitrary Binary Operation}\\
    & \mid & \unop e & \textit{Arbitrary Unary Operation}\\
    \multicolumn{3}{l}{\Table} \\
    t & ::= & \left\{\begin{array}{l}
    t.\id = v; \\
    t.\keys = \List[h.f]; \\
    t.\actions = \List[\lambda \vec x. c]\\
    \end{array}\right\} & \textit{Table}\\
    \multicolumn{3}{l}{\BExpr} \\
    b & ::= & \FALSE  & \textit{Falsity}\\
      & \mid & b \Rightarrow b & \textit{Implication} \\
      & \mid & e = e & \textit{Bitvector Equality}\\
    & \mid & \forall \rho \in t.~b & \textit{Row Universality}\\
    \multicolumn{3}{l}{\Cmd} \\
    c & ::=  & h.f := e & \textit{Assignment}\\
      & \mid & \assert b & \textit{Assertion}\\
      & \mid & \assume b & \textit{Assumption}\\
      & \mid & \choose \rho t & \textit{Selection} \\
      & \mid & t.\apply() & \textit{Table Application}\\
      & \mid & c;c & \textit{Sequential Composition} \\
    & \mid & c \choice c & \textit{Nondeterministic Choice}\\
    \multicolumn{3}{l}{
      \begin{array}{ll}
        x \in \Var & h \in \Hdr\\
        n,w,i \in \mathbb{N} & f \in \Field
    \end{array}} & \textit{Sets }\\
  \end{array}
\]
\caption{Grammar of Dataplane Programs}
\label{fig:grammar}
\end{figure}

@@

\paragraph{Commands}
Data plane programs $c \in \mathsf{Cmd}$ can be assignments ($h.f := e$),
assumptions ($\assume b$), assertions ($\assert b$), table application
($t.\apply()$), nondeterministic selection ($\choose \rho t$), sequential composition ($c_1;c_2$ for two commands $c_1$ and
$c_2$), and nondeterministic choice ($c_1 \choice c_2$ for two commands $c_1$
and $c_2$).

@@

\paragraph{Row Models}
We also need models $\mu \in \Model$ to indicate which symbolic rows corresponds
to which concrete rows, i.e. $\Model$ is a map from symbolic rows to concrete
rows; write $\mu(\rho) = r$ for a model $\mu$, a symbolic row $\rho$, and a
concrete row $r$.

@@

\paragraph{Command Evaluation}
Commands are evaluated similar to Avenir in Figure~\ref{fig:com-sem}. The
notable differences are the inclusion of assertions, the use of a binary
nondeterminism operator, and a cleaner representation of table application.

\begin{figure}[htp]
  \[\begin{array}{l >{\triangleq}cl}
  \multicolumn{3}{l}{\denote{c}^\sigma : \Model \to \Pkt \to \mathcal{P}(\Model \times \Pkt)_\error }\\
    \denote{h.f := e}^\sigma~\mu~\pkt && \{(\mu, \pkt\{h \mapsto \edenote{e}~\pkt~\mu\})\} \\
    \denote{\assume b}^\sigma~\mu~\pkt &&
    \begin{cases}
      \{(\mu,\pkt)\} & \pkt,\mu \models_\sigma b \\
      \{\} & \textit{otherwise} \\
    \end{cases}\\
    \denote{\assert b}^\sigma~\mu~\pkt &&
    \begin{cases}
      \{(\mu,\pkt)\} & \pkt,\mu \models_\sigma b \\
      \error & \textit{otherwise}
    \end{cases} \\
    \denote{\choose \rho t}^\sigma~\mu~\pkt &&
    \bigcup_{r \in \sigma(t)}\{(\pkt,\mu\{\rho \mapsto r\})\}
    \\
    \denote{c_1;c_2}^\sigma~\mu~\pkt&&
    \bigcup_{(\mu',\pkt') \in \denote{c_1}^\sigma\,\mu\,\pkt}\denote{c_2}^\sigma~\mu'~\pkt' \\
    \denote{c_1 \choice c_2}^\sigma~\mu~\pkt &&
    \denote{c_1}^\sigma~\mu~\pkt \cup \denote{c_2}^\sigma~\mu~\pkt \\
    \denote{t.\apply()}^\sigma~\mu~\pkt &&
    \begin{cases}
      \denote{c[\vec d/\vec x]}^\sigma~\mu~\pkt,
      & \text{where}~\lambda \vec x.~c = r.\action, \vec d = r.\data \\
      & \text{where}~r = \argmin_i\{\pkt,\mu \models_\sigma \sigma(t.\id)[i].\matches = t.\keys \} \\
      \{(\mu,\pkt)\}, & \mathit{otherwise}
    \end{cases}
  \end{array}\]
  \caption{Semantics of Commands}
  \label{fig:com-sem}
  \todo[inline]{$\Model$ then $\Pkt$ everywhere.}
\end{figure}

@@

\begin{figure}
  \[\begin{array}{r>{\triangleq}cl}
  (h.f := e)[\sigma] && h.f := e   \\
  (\assume b)[\sigma] && \assume b \\
  (\assert b)[\sigma] && \assert b \\
  (c_1;c_2)[\sigma] && c_1[\sigma];c_2[\sigma] \\
  (c_1 \choice c_2)[\sigma] && c_1[\sigma] \choice c_2[\sigma] \\
  (\choose \rho t)[\sigma] && \textrm{\textcolor{red}{unnecessary, but similar to t.apply}}\\
  (t.\apply())[\sigma] &&
  \begin{array}[t]{l}
    \{\\
    \quad \assume {r_1.\matches = t.\keys};\\
    \quad t.\action[r_1.\action](r_1.\data)\\
    \}~\choice\\
    \multicolumn{1}{c}{\vdots}\\
    \choice~\{\\
    \quad \assume {r_n.\matches = t.\keys};\\
    \quad t.\action[r_n.\action](r_n.\data)\\
    \}\\
    \textrm{where}~[r_1;\ldots;r_n] = \sigma(t.\id)
  \end{array}
  \end{array}
  \]
  \caption{Table Elimination}
  \label{fig:table-elim}
\end{figure}

@@

\begin{figure}[htp]
  \[\begin{array}{r >{\triangleq}cl}
  (h.f := e)[?] && h.f := e \\
  (\assume b)[?] && \assume b \\
  (\assert b)[?] && \assert b\\
  (c_1;c_2)[?] && c_1[?];c_2[?]\\
  (c_1 \choice c_2)[?] && c_1[?] \choice c_2[?]\\
  (\choose \rho t)[?] &&
  \textrm{\textcolor{red}{unnecessary}} \\
  (t.\apply())[?] &&
  \begin{array}[t]{l} \displaystyle
    \{ \\
    \quad \choose \rho t; \\
    \quad \assume \rho.\matches = t.\keys; \\
    \quad \displaystyle \bigchoice_{i=0}^{|t.\action| -1}
    \left\{\begin{array}{l}
    \assume \rho.\action = i;\\
    t.\action[i](\rho.\data)
    \end{array}\right\}\\
    \} \choice \{ \\
    \quad \assume \forall \rho \in t. \rho.\matches \neq t.\keys; \\
    \quad \SKIP \\
    \quad \}\\
    \}
  \end{array}
  \end{array}
  \]
  \caption{Symbolic Table instrumentation}
  \label{fig:table-instrument}
\end{figure}

@@

      Now let's consider the first branch, considering each command sequentially.
      \[\begin{array}{>{\displaystyle}ll}
      \denote{\choose \rho t; \assume \rho.\matches = t.\keys}^\sigma~\mu~\pkt \\
      = \bigcup_{r \in \sigma(t)}\denote{\assume\rho.\matches = t.\keys}^\sigma~\mu\{\rho \mapsto r\}~\pkt
      & \textrm{by definition of }\denote{-}\\
      = \{(\mu\{\rho\mapsto r\},\pkt)
      \mid r \in \sigma(t), \pkt,\mu\{\rho\mapsto r\} \models_\sigma \rho.\matches = t.\keys
      \}
      & \textrm{by definitions and set theory} \\
      = \{(\mu\{\rho \mapsto r\}, \pkt)
      \mid r \in \sigma(t), \pkt,\mu \models_\sigma r.\matches = t.\keys\}
      & \textrm{by definitions \textcolor{red}{TODO Lemma}} \\
      = \{(\mu\{\rho \mapsto r_i\}, \pkt)\}
      & \textrm{by \textsc{Hit} and \textsc{NoShadow}} \\
      \end{array}\]

      Now by the sequential composition rule, we can determine the result of the
      first branch by computing as follows:
      \[\begin{array}{>{\displaystyle}ll}
      \denote{\bigchoice_{i=0}^{|t.\action| -1}
        \left\{\begin{array}{l}
        \assume \rho.\action = i;\\
        t.\actions[i](\rho.\data)
        \end{array}\right\}
      }^\sigma~(\mu\{\rho \mapsto r_i\})~\pkt \\
      = \denote{\denote{c[r_i.\data/\vec x]}^\sigma~(\mu\{\rho \mapsto r\})~\pkt}\\
      & \textrm{skipping lots of definition steps} \\
      \end{array}
      \]
      \todo[inline]{Need Scoping Rules for $\rho$}

>>>>

% commands
\newcommand{\assert}{\mathop{\mathsf{assert}}}
\newcommand{\assume}{\mathop{\mathsf{assume}}}
\newcommand{\apply}{\mathsf{apply}}
\newcommand{\choiceop}{\rotatebox[origin=c]{90}{$\sqsubset\!\!\!\sqsupset$}}
\newcommand{\choice}{\mathbin{\choiceop}}
\DeclareMathOperator*{\bigchoice}{\scalerel*{\choiceop}{\sum}}
\renewcommand{\choose}[2]{\mathop{\mathsf{choose}~#1~\mathsf{from}~#2~\mathsf{in}}}
\newcommand{\SKIP}{\mathsf{skip}}

@@

\begin{figure}[htp]
  \[\begin{array}{lcll}
    \multicolumn{3}{l}{\Value} \\
    v & ::= & [n]_w  & \textit{Bitvector Literal}\\
    \multicolumn{3}{l}{\SymbRow} \\
    \rho & ::= & \left\{
    \begin{array}{l}
      \matches = \List[x];\\
      \action = x;\\
      \data = \List[x]
    \end{array}\right\} & \textit{Symbolic Row}\\
    \multicolumn{3}{l}{\BVExpr} \\
    e & ::= & v & \textit{Value}\\
      & \mid & h.f & \textit{Header Access} \\
      & \mid & \rho.\matches[i] & \textit{Symbolic Match} \\
      & \mid & \rho.\action & \textit{Symbolic Action}\\
      & \mid & \rho.\data[i] & \textit{Symbolic Action Data}\\
      & \mid & x & \textit{Variable}\\
      & \mid & e \binop e & \textit{Arbitrary Binary Operation}\\
    & \mid & \unop e & \textit{Arbitrary Unary Operation}\\
    \multicolumn{3}{l}{\Table} \\
    t & ::= & \left\{\begin{array}{l}
    t.\id = v; \\
    t.\keys = \List[h.f]; \\
    t.\actions = \List[\lambda \vec x. c]\\
    \end{array}\right\} & \textit{Table}\\
    \multicolumn{3}{l}{\BExpr} \\
    b & ::= & \FALSE  & \textit{Falsity}\\
      & \mid & b \Rightarrow b & \textit{Implication} \\
      & \mid & e = e & \textit{Bitvector Equality}\\
    & \mid & \forall \rho \in t.~b & \textit{Row Universality}\\
    \multicolumn{3}{l}{\Cmd} \\
    c & ::=  & h.f := e & \textit{Assignment}\\
      & \mid & \assert b & \textit{Assertion}\\
      & \mid & \assume b & \textit{Assumption}\\
      & \mid & t.\apply() & \textit{Table Application}\\
      & \mid & c;c & \textit{Sequential Composition} \\
    & \mid & c \choice c & \textit{Nondeterministic Choice}\\
    & \mid & \choose \rho t c & \textit{Selection} \\
    \multicolumn{3}{l}{
      \begin{array}{ll}
        x \in \Var & h \in \Hdr\\
        n,w,i \in \mathbb{N} & f \in \Field
    \end{array}} & \textit{Sets }\\
  \end{array}
\]
\caption{Grammar of Dataplane Programs}
\label{fig:grammar}
\end{figure}

@@

\paragraph{Commands}
Data plane programs $c \in \mathsf{Cmd}$ can be assignments ($h.f := e$),
assumptions ($\assume b$), assertions ($\assert b$), table application
($t.\apply()$), lexically-scoped nondeterministic selection ($\choose \rho t c$), sequential composition ($c_1;c_2$ for two commands $c_1$ and
$c_2$), and nondeterministic choice ($c_1 \choice c_2$ for two commands $c_1$
and $c_2$).

@@

\paragraph{Row Models}
We also need models $\mu \in \Model$ to indicate which symbolic rows corresponds
to which concrete rows, i.e. $\Model = \SymbRow \rightharpoonup \Row$. We write
$\mu\{\rho \mapsto r\}$ to extend $\mu$ by mapping $\rho$ to $r$, and write $\mu
\setminus \rho$, to remove $\rho$ from the map.

@@

\paragraph{Command Evaluation}
Commands are evaluated similar to Avenir in Figure~\ref{fig:com-sem}. The
notable differences are the inclusion of assertions, the nondeterminism
operators, and a cleaner representation of table application.

\begin{figure}[htp]
  \[\begin{array}{l >{\triangleq}cl}
  \multicolumn{3}{l}{\denote{c}^\sigma : \Model \to \Pkt \to \mathcal{P}(\Model \times \Pkt)_\error }\\
    \denote{h.f := e}^\sigma~\mu~\pkt && \{(\mu, \pkt\{h \mapsto \edenote{e}~\pkt~\mu\})\} \\
    \denote{\assume b}^\sigma~\mu~\pkt &&
    \begin{cases}
      \{(\mu,\pkt)\} & \pkt,\mu \models_\sigma b \\
      \{\} & \textit{otherwise} \\
    \end{cases}\\
    \denote{\assert b}^\sigma~\mu~\pkt &&
    \begin{cases}
      \{(\mu,\pkt)\} & \pkt,\mu \models_\sigma b \\
      \error & \textit{otherwise}
    \end{cases} \\
    \denote{\choose \rho t c}^\sigma~\mu~\pkt &&
    \{(\mu'\setminus \rho,\pkt')
    \mid (\mu',\pkt') \in \denote{c}^\sigma\,(\mu\{\rho\mapsto r\})\,\pkt, r \in \sigma(t)\}
    \\
    \denote{c_1;c_2}^\sigma~\mu~\pkt&&
    \bigcup_{(\mu',\pkt') \in \denote{c_1}^\sigma\,\mu\,\pkt}\denote{c_2}^\sigma~\mu'~\pkt' \\
    \denote{c_1 \choice c_2}^\sigma~\mu~\pkt &&
    \denote{c_1}^\sigma~\mu~\pkt \cup \denote{c_2}^\sigma~\mu~\pkt \\
    \denote{t.\apply()}^\sigma~\mu~\pkt &&
    \begin{cases}
      \denote{c[\vec d/\vec x]}^\sigma~\mu~\pkt,
      & \text{where}~\lambda \vec x.~c = r.\action, \vec d = r.\data \\
      & \text{where}~r = \argmin_i\{\pkt,\mu \models_\sigma b_i\} \\
      & \text{where}~b_i = \sigma(t.\id)[i].\matches = t.\keys \\
      \{(\mu,\pkt)\}, & \mathit{otherwise}
    \end{cases}
  \end{array}\]
  \caption{Semantics of Commands}
  \label{fig:com-sem}
  \todo[inline]{$\Model$ then $\Pkt$ everywhere.}
\end{figure}

@@

\begin{figure}
  \[\begin{array}{r>{\triangleq}cl}
  (h.f := e)[\sigma] && h.f := e   \\
  (\assume b)[\sigma] && \assume b \\
  (\assert b)[\sigma] && \assert b \\
  (c_1;c_2)[\sigma] && c_1[\sigma];c_2[\sigma] \\
  (c_1 \choice c_2)[\sigma] && c_1[\sigma] \choice c_2[\sigma] \\
  (\choose \rho t c)[\sigma] && \textrm{\textcolor{red}{unnecessary, but similar to t.apply}}\\
  (t.\apply())[\sigma] &&
  \begin{array}[t]{l}
    \{\\
    \quad \assume {r_1.\matches = t.\keys};\\
    \quad t.\action[r_1.\action](r_1.\data)\\
    \}~\choice\\
    \multicolumn{1}{c}{\vdots}\\
    \choice~\{\\
    \quad \assume {r_n.\matches = t.\keys};\\
    \quad t.\action[r_n.\action](r_n.\data)\\
    \}\\
    \textrm{where}~[r_1;\ldots;r_n] = \sigma(t.\id)
  \end{array}
  \end{array}
  \]
  \caption{Table Elimination}
  \label{fig:table-elim}
\end{figure}

@@

\begin{figure}[htp]
  \[\begin{array}{r >{\triangleq}cl}
  (h.f := e)[?] && h.f := e \\
  (\assume b)[?] && \assume b \\
  (\assert b)[?] && \assert b\\
  (c_1;c_2)[?] && c_1[?];c_2[?]\\
  (c_1 \choice c_2)[?] && c_1[?] \choice c_2[?]\\
  (\choose \rho t c)[?] &&
  \textrm{\textcolor{red}{unnecessary}} \\
  (t.\apply())[?] &&
  \begin{array}[t]{l} \displaystyle
    \{ \\
    \quad \choose \rho t \;\{ \\
    \qquad \assume \rho.\matches = t.\keys; \\
    \qquad \displaystyle \bigchoice_{i=0}^{|t.\action| -1}
    \left\{\begin{array}{l}
    \assume \rho.\action = i;\\
    t.\action[i](\rho.\data)
    \end{array}\right\}\\
    \quad \} \\
    \} \choice \{ \\
    \quad \assume \forall \rho \in t. \rho.\matches \neq t.\keys; \\
    \quad \SKIP \\
    \quad \}\\
    \}
  \end{array}
  \end{array}
  \]
  \caption{Symbolic Table instrumentation}
  \label{fig:table-instrument}
\end{figure}

@@

      Now let's consider the first branch, i.e.
      \[
      \denote{\choose \rho t \left\{
        \begin{array}{>{\displaystyle}l}
          \assume \rho.\matches = t.\keys;\\
          \bigchoice_{i=0}^{|t.\action| -1}
          \left\{\begin{array}{l}
          \assume \rho.\action = i;\\
          t.\actions[i](\rho.\data)
          \end{array}\right\}
        \end{array}\right\}
      }^\sigma~\mu~\pkt\]

      Compute the following inner result, by definitions and set theory
      \[
      \begin{array}{>{\displaystyle}ll}
      \bigcup_{r \in \sigma(t)}\denote{\begin{array}{>{\displaystyle}l}
          \assume \rho.\matches = t.\keys;\\
          \bigchoice_{i=0}^{|t.\action| -1}
          \left\{\begin{array}{l}
          \assume \rho.\action = i;\\
          t.\actions[i](\rho.\data)
          \end{array}\right\}
      \end{array}}^\sigma~\mu\{\rho \mapsto r\}~\pkt\\
      = \bigcup_{
        \substack{r \in \sigma(t) \\
          (\mu',\pkt') \in \denote{\assume \rho.\matches = t.\keys}^\sigma\,(\mu\{\rho \mapsto r\})\,\pkt}}
      \bigcup_{i=0}^{|t.\action| -1}\denote{
       \begin{array}{l}
          \assume \rho.\action = i;\\
          t.\actions[i](\rho.\data)
      \end{array}}^\sigma~\mu'~\pkt'

      \\\\
      (\textrm{by \textsc{Hit}, \textsc{NoShadow}, and the definition of }\denote{-}) \\
      \\
      = \bigcup_{i=0}^{|t.\action| -1}\denote{
       \begin{array}{l}
          \assume \rho.\action = i;\\
          t.\actions[i](\rho.\data)
      \end{array}}^\sigma~(\mu\{\rho \mapsto r_i\})~\pkt \\
      = \bigcup_{j=0}^{|t.\action| -1}\denote{
       \begin{array}{l}
          \assume r_i.\action = j;\\
          t.\actions[j](r_i.\data)
      \end{array}}^\sigma~(\mu\{\rho \mapsto r_i\})~\pkt \\
      \textrm{\textcolor{red}{TODO LEMMA}} \\
      = \denote{t.\actions[r_i.\action](r_i.\data)}^\sigma~(\mu\{\rho \mapsto r_i\})~\pkt \\
      = \denote{c[r_i.\data/\vec x]}^\sigma~(\mu\{\rho \mapsto r_i\})~\pkt
      \end{array}\]

      Since $\rho$ is fresh in $c$, and consequently in $c[r_i.\data/\vec x]$,
      then by \textcolor{red}{Lemma},
      \[\{(\mu'\setminus \rho,\pkt')
      \mid (\mu', \pkt') \in \denote{c[r_i.\data/\vec x]}^\sigma~(\mu\{\rho \mapsto r_i\})~\pkt\}
       = \denote{c[r_i.\data/\vec x]}^\sigma~\mu~\pkt
       \]
       \todo[inline]{Relies on the fact that $\rho \not \in \mu$, or that $\rho$ is globally fresh}

       The result follows.
